---
title: "Japan, Covid-19 situation by prefecture"
author: "Mitsuo Shiota"
date: "2021-03-31"
output: 
  github_document:
    toc: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updated: `r Sys.Date()`

## Summary

I added Japan page to [my Shiny App](https://mitsuoxv.shinyapps.io/covid/) on March 31, 2021. I use data of 47 prefectures from [NHK site ](https://www3.nhk.or.jp/news/special/coronavirus/data/).

```{r library, include=FALSE}
library(tidyverse)
library(lubridate)

library(NipponMap)
library(sf)

library(slider)
```

## Read data from NHK site

```{r read_data_from_NHK, message=FALSE}

data_raw <- read_csv("https://www3.nhk.or.jp/n-data/opendata/coronavirus/nhk_news_covid19_prefectures_daily_data.csv")
names(data_raw) <- c("publish_date", "code", "prefecture",
                     "new_conf", "cum_conf", "new_deaths", "cum_deaths")

data_japan <- data_raw %>% 
  mutate(publish_date = ymd(publish_date))

```

## Add population column

I download population by prefecture data from [e-stat page](https://www.e-stat.go.jp/stat-search/database?page=1&layout=datalist&toukei=00200524&bunya_l=02&tstat=000000090001&cycle=0&tclass1=000001136886&statdisp_id=0003412316&tclass2val=0). The data are estimated as of October 1, 2019.

```{r population, include=FALSE}

pop <- read_csv("data-raw/FEH_00200524_210331081757.csv", 
                locale = locale(encoding = "SHIFT-JIS"), 
                skip = 12)

pop <- pop[, c(3, 6)]
names(pop) <- c("code", "population")

pref_eng <- read_csv("data-raw/pref_eng.csv")

pop <- bind_cols(pop, pref_eng)

pop <- pop %>% 
  mutate(
    code = code %>% str_sub(end = 2L),
    population = population * 1000
    )

data_japan <- data_japan %>% 
  select(-prefecture) %>% 
  left_join(pop, by = "code")

prefectures_map <-
  st_read(system.file("shapes/jpn.shp", package = "NipponMap"))

in_japan <- data_japan %>%
  pivot_longer(!c(publish_date, code, prefecture, population),
               names_to = "concept", values_to = "value") %>% 
  group_by(code, concept) %>% 
  mutate(
    value_ma = slide_dbl(value, mean, na.rm = TRUE,
                       .before = 6, .complete = TRUE),
    value_per1m = value / population * 1e6,
    value_ma_per1m = value_ma / population * 1e6
      ) %>% 
  ungroup()

prefecture_menu <- unique(in_japan$prefecture)

# make a list
japan <- list(
  map_df = prefectures_map,
  area_df = in_japan,
  area_menu = prefecture_menu
)

usethis::use_data(japan, overwrite = TRUE)

```


```{r map_draw_prepare, include=FALSE}

# read map data
prefectures_map <-
  st_read(system.file("shapes/jpn.shp", package = "NipponMap"))

draw_map_japan <- function(map_df, df, var_str) {
  map_df %>%
    left_join(df, by = c(jiscode = "code")) %>%
    ggplot() +
    geom_sf(aes(fill = .data[[var_str]]), color = "white") +
    scale_fill_gradient2(low = "#559999", mid = "grey90", high = "#BB650B",
                         midpoint = median(df[[var_str]])) +
    theme_void()
}

```

## Fastest spreading prefecutures in the last 14 days

Highest "increase_in_14_days", which is an increase in cumulative confirmed cases in the last 14 days, are:

```{r spreading, echo=FALSE}
prefectures_speed <- data_japan %>% 
  group_by(code) %>% 
  filter(publish_date > max(publish_date) - 14) %>% 
  summarize(increase_in_14_days = sum(new_conf), .groups = "drop_last")

draw_map_japan(prefectures_map, prefectures_speed, "increase_in_14_days")

```

Above calculation might be unfair to populous prefectures. Highest "increase_std_in_14_days", which is an increase in cumulative confirmed cases in the last 14 days per 1 million population, are:

```{r standardized, echo=FALSE}
prefectures_speed_std <- data_japan %>% 
  group_by(code) %>% 
  filter(publish_date > max(publish_date) - 14) %>% 
  summarize(increase_std_in_14_days = sum(new_conf * 1000000 / population), .groups = "drop_last")

draw_map_japan(prefectures_map, prefectures_speed_std, "increase_std_in_14_days")

```


## Highest fatality rate prefectures

Among prefectures with more than 10 cumulative deaths, highest "fatality_rate", which is cumulative deaths per 100 cumulative confirmed cases, are:

```{r fatality_rates, echo=FALSE}
prefectures_fatality <- data_japan %>% 
  filter(
    cum_deaths > 10,
    publish_date == max(publish_date)
    ) %>% 
  mutate(fatality_rate = cum_deaths * 100 / cum_conf)

draw_map_japan(prefectures_map, prefectures_fatality, "fatality_rate")

```

## Highest deaths per population prefectures

Highest "deaths_per_1m", which is cumulative deaths per 1 million population, are:

```{r deaths_per_population, echo=FALSE}
prefectures_deaths <- data_japan %>% 
  filter(publish_date == max(publish_date)) %>% 
  mutate(
    pop_mil = population / 1000000,
    deaths_per_1m = cum_deaths / pop_mil
    )

draw_map_japan(prefectures_map, prefectures_deaths, "deaths_per_1m")

```

EOL
