---
title: "Japan, weekly confirmed Covid-19 cases per hospital"
author: "Mitsuo Shiota"
date: "2023-05-20"
format: gfm
execute:
  echo: false
---

Updated: `r Sys.Date()`

## Covid-19 was reclassified as tier 5 on May 8, 2023 in Japan

There are no daily data available on Covid-19 since May 8, 2023 in Japan. Instead, MHLW (Ministry of Health, Labour and Welfare) publishes weekly data in [its web site](https://www.mhlw.go.jp/stf/seisakunitsuite/bunya/0000121431_00086.html).

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(readxl)

library(pdftools)

theme_set(theme_light())
```

## Read data from MHLW site

```{r}
#| label: pref_eng

pref_eng <- read_csv("data-raw/pref_eng.csv") |> 
  mutate(code = sprintf("%02.0f", 0:47))
```

```{r}
#| label: read_data_from_MHLW
#| message: false

end_row_num <- 36
url_excel <- "data-raw/001098622.xlsx"

conf_num <- read_excel(url_excel,
                       range = paste0("A5:AX", end_row_num)) |>
  set_names("start", "end", pref_eng$prefecture) |>
  pivot_longer(!start:end, names_to = "prefecture", values_to = "value") |>
  mutate(
    start = as.Date(start),
    concept = "new_conf_wk"
    ) |> 
  select(publish_date = start, prefecture, value)

conf_per_hospital <- read_excel(url_excel,
                   range = paste0("A", end_row_num + 6, ":AX", end_row_num * 2 + 1)) |> 
  set_names("start", "end", pref_eng$prefecture) |>
  pivot_longer(!start:end, names_to = "prefecture", 
               values_to = "value_per_hospital") |> 
  select(value_per_hospital)

in_japan_wk <- bind_cols(conf_num, conf_per_hospital) |> 
  left_join(pref_eng, by = "prefecture")
```

```{r}
#| label: prepare_for_pdf

# Split by line
make_lines <- function(a_list) {
  a_list |> 
    str_split("\n") |> 
    unlist()
}

# read character vector into a tibble, split by >= 5 spaces
read_chr_vec <- function(chr_vec, skip, expect_col_num) {
  
  chr_vec[(skip + 1):length(chr_vec)] |> 
    str_split("\\s{5,}") |> 
    unlist() |> 
    matrix(ncol = expect_col_num, byrow = TRUE) |> 
    as_tibble(.name_repair = "minimal")
}

# read pdf and return df_table
read_pdf <- function(url_pdf) {
  # Extract text
  texts <- pdf_text(url_pdf)
  
  # Split into lines
  lines <- texts  %>% 
    make_lines()
  
  # Table
  table_start <- str_which(lines, "新型コロナウイルス感染症（COVID-19）定点当たり報告数・都道府県別")
  
  table_end <- str_which(lines, "新型コロナウイルス感染症（COVID-19）定点当たり報告数推移")
  
  lines_table <- lines[table_start:(table_end - 2)]
    
  df_table <- read_chr_vec(lines_table, skip = 5, expect_col_num = 3) |> 
    set_names("prefecture_kj", "value", "value_per_hospital") |> 
    mutate(across(value:value_per_hospital, parse_number))
  
  date_line <- lines_table[2]
  
  py <- str_extract(date_line, "(\\d+)年", group = 1) |> as.numeric()
  pm <- str_extract(date_line, "(\\d+)月", group = 1) |> as.numeric()
  pd <- str_extract(date_line, "(\\d+)日", group = 1) |> as.numeric()
  
  # add publish_date
  bind_cols(df_table[c(48, 1:47), ], pref_eng) |> 
    mutate(publish_date = make_date(py, pm, pd)) |> 
    select(-prefecture_kj)
}
```

```{r}
#| label: read_from_pdf

urls <- paste0("https://www.mhlw.go.jp/content/",
               c("001100553", "001102876", "001103206", "001105457",
                 "001108590", "001111891", "001115055", "001118343",
                 "001120797", "001122882", "001126858", "001131133",
                 "001133321", "001135122", "001138245", "001140810",
                 "001144212", "001146197", "001148508", "001151300",
                 "001154253", "001155887", "001158575", "001161146",
                 "001163905", "001165794", "001168343", "001170443",
                 "001173342", "001176400", "001178927", "001181783",
                 "001185610", "001187734", "001188996", "001193285",
                 "001197418", "001202247", "001206921", "001210301",
                 "001214098", "001216718", "001220751", "001225957",
                 "001231623"),
               ".pdf")

df_table <- map(urls, read_pdf) |> 
  list_rbind()

# add pdf data
in_japan_wk <- bind_rows(in_japan_wk, df_table)
```

## 2 csv files

I make 2 csv files.

- japan_wk_cases.csv
- japan_wk_cases_per_hospital.csv

```{r}
#| label: write_csv_files

in_japan_wk |> 
  select(code, prefecture, publish_date, value) |> 
  pivot_wider(names_from = publish_date, values_from = value) |> 
  write_csv("japan_wk_cases.csv")

in_japan_wk |> 
  select(code, prefecture, publish_date, value_per_hospital) |> 
  pivot_wider(names_from = publish_date, values_from = value_per_hospital) |> 
  write_csv("japan_wk_cases_per_hospital.csv")
```

## Weekly confirmed cases per hospital

```{r}
#| label: plot_line_chart

in_japan_wk |> 
  filter(prefecture != "Total") |> 
  ggplot(aes(publish_date, value_per_hospital)) +
  geom_line(aes(color = prefecture), show.legend = FALSE) +
  geom_line(data = in_japan_wk |> filter(prefecture == "Total"),
            color = "black", linewidth = 1) +
  labs(x = NULL, y = "Cases per hospital",
   title = paste0("Weekly confirmed cases; latest is the week starting on ", max(in_japan_wk$publish_date)),
       subtitle = "47 prefectures are colored lines, and the national average is a black line",
       caption = "Source: Ministry of Health, Labour and Welfare")
```

For more flexible data presentations, please check [my Shiny app](https://mitsuoxv.shinyapps.io/covid/).

```{r}
#| label: japan_wk
#| include: false

# make a list
japan_wk <- list(
#  map_df = prefectures_map,
  area_df = in_japan_wk,
  area_menu = pref_eng$prefecture
)

usethis::use_data(japan_wk, overwrite = TRUE)
```

EOL
