---
title: "Japan, Covid-19 situation by prefecture"
author: "Mitsuo Shiota"
date: "2021-03-31"
output: 
  github_document:
    toc: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updated: `r Sys.Date()`

## Summary

I added Japan page to [my Shiny App](https://mitsuoxv.shinyapps.io/covid/) on March 31, 2021. I used data of 47 prefectures from [NHK site ](https://www3.nhk.or.jp/news/special/coronavirus/data/).

On September 28, 2022, NHK stopped to collect data by itself, as MHLW (Ministry of Health, Labour  and Welfare) began to publish [the data](https://www.mhlw.go.jp/stf/covid-19/open-data.html). So, I changed the source from NHK to MHLW.

```{r library, include=FALSE}
library(tidyverse)
library(readxl)
library(lubridate)

library(NipponMap)
library(sf)

library(tqr)
```

## Read data from MHLW site

```{r read_data_from_MHLW, message=FALSE}
conf <- read_csv("https://covid19.mhlw.go.jp/public/opendata/newly_confirmed_cases_daily.csv") %>% 
  pivot_longer(-Date, names_to = "prefecture", values_to = "new_conf") %>% 
  filter(prefecture != "ALL") %>% 
  mutate(publish_date = ymd(Date)) %>% 
  group_by(prefecture) %>% 
  mutate(cum_conf = cumsum(new_conf)) %>% 
  ungroup() %>% 
  select(publish_date, prefecture, new_conf, cum_conf)

deaths <- read_csv("https://covid19.mhlw.go.jp/public/opendata/deaths_cumulative_daily.csv") %>% 
  pivot_longer(-Date, names_to = "prefecture", values_to = "cum_deaths") %>% 
  filter(prefecture != "ALL") %>% 
  mutate(publish_date = ymd(Date)) %>% 
  group_by(prefecture) %>% 
  mutate(
    cum_deaths_lag = lag(cum_deaths),
    new_deaths = cum_deaths - cum_deaths_lag
    ) %>% 
  ungroup() %>% 
  select(publish_date, prefecture, new_deaths, cum_deaths)

prefectures <- tibble(
  prefecture = conf %>% 
    filter(publish_date == "2020-01-16") %>% 
    pull(prefecture),
  code = str_c("0", 1:47) %>% 
    str_sub(start = -2L)
)

data_japan <- conf %>% 
  left_join(deaths, by = c("publish_date", "prefecture")) %>% 
  left_join(prefectures, by = "prefecture")

```

## Add population column

I download population by prefecture data from [the Statistics Bureau, Ministry of Internal Affairs and Communications](https://www.stat.go.jp/data/nihon/02.htm). The data are estimated as of October 1, 2020.

```{r population, include=FALSE}

pop <- read_excel("data-raw/n220200200.xlsx", skip = 5)

pop <- pop[2:49, 1:2] %>% 
  setNames(c("prefecture", "population")) %>% 
  mutate(
    code = sprintf("%02.0f", 0:47),
    population = population * 1000
    ) %>% 
  select(-prefecture)

pref_eng <- read_csv("data-raw/pref_eng.csv")

pop <- bind_cols(pop, pref_eng)

data_japan <- data_japan %>% 
  select(-prefecture) %>% 
  left_join(pop, by = "code")

prefectures_map <-
  st_read(system.file("shapes/jpn.shp", package = "NipponMap"))

# inset Okinawa
prefectures_map$geometry[[47]] <- prefectures_map$geometry[[47]] + c(7, 14)

in_japan <- data_japan %>%
  pivot_longer(!c(publish_date, code, prefecture, population),
               names_to = "concept", values_to = "value") %>% 
  group_by(code, concept) %>% 
  mutate(
    value_ma = moving_average(value, n = 7, na.rm = TRUE),
    value_per1m = value / population * 1e6,
    value_ma_per1m = value_ma / population * 1e6
      ) %>% 
  ungroup()

prefecture_menu <- unique(in_japan$prefecture)

# make a list
japan <- list(
  map_df = prefectures_map,
  area_df = in_japan,
  area_menu = prefecture_menu
)

usethis::use_data(japan, overwrite = TRUE)

```


```{r map_draw_prepare, include=FALSE}

draw_map_japan <- function(map_df, df, var) {
  lines <- cbind(c(132, 135, 137, 137), c(38, 38, 40, 43)) |> 
    st_linestring()
  
  map_df %>%
    left_join(df, by = c(jiscode = "code")) %>%
    ggplot() +
    geom_sf(aes(fill = {{ var }}), color = "white") +
    geom_sf(data = lines, color = "gray80") +
    xlim(130, 149) +
    ylim(31, 45) +
    scale_fill_gradient2(low = "#559999", mid = "grey90", high = "#BB650B",
                         midpoint = median(df %>% pull({{ var }}))) +
    theme_void()
}

```

## Fastest spreading prefecutures in the last 14 days

Highest "increase_in_14_days", which is an increase in cumulative confirmed cases in the last 14 days, are:

```{r spreading, echo=FALSE}
prefectures_speed <- data_japan %>% 
  group_by(code) %>% 
  filter(publish_date > max(publish_date) - 14) %>% 
  summarize(increase_in_14_days = sum(new_conf), .groups = "drop_last")

draw_map_japan(prefectures_map, prefectures_speed, increase_in_14_days)

```

Above calculation might be unfair to populous prefectures. Highest "increase_std_in_14_days", which is an increase in cumulative confirmed cases in the last 14 days per 1 million population, are:

```{r standardized, echo=FALSE}
prefectures_speed_std <- data_japan %>% 
  group_by(code) %>% 
  filter(publish_date > max(publish_date) - 14) %>% 
  summarize(increase_std_in_14_days = sum(new_conf * 1000000 / population), .groups = "drop_last")

draw_map_japan(prefectures_map, prefectures_speed_std, increase_std_in_14_days)

```


## Highest fatality rate prefectures

Among prefectures with more than 10 cumulative deaths, highest "fatality_rate", which is cumulative deaths per 100 cumulative confirmed cases, are:

```{r fatality_rates, echo=FALSE}
prefectures_fatality <- data_japan %>% 
  filter(
    cum_deaths > 10,
    publish_date == max(publish_date)
    ) %>% 
  mutate(fatality_rate = cum_deaths * 100 / cum_conf)

draw_map_japan(prefectures_map, prefectures_fatality, fatality_rate)

```

## Highest deaths per population prefectures

Highest "deaths_per_1m", which is cumulative deaths per 1 million population, are:

```{r deaths_per_population, echo=FALSE}
prefectures_deaths <- data_japan %>% 
  filter(publish_date == max(publish_date)) %>% 
  mutate(
    pop_mil = population / 1000000,
    deaths_per_1m = cum_deaths / pop_mil
    )

draw_map_japan(prefectures_map, prefectures_deaths, deaths_per_1m)

```

EOL
