---
title: "USA, Covid-19 situation by state"
author: "Mitsuo Shiota"
date: "2020-05-26"
output: 
  github_document:
    toc: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Updated: `r Sys.Date()`

## Summary

I added the United States page to [my Shiny App](https://mitsuoxv.shinyapps.io/covid/) on May 25, 2020. I use data of 50 states and 1 federal district (District of Columbia) from [USAFacts page](https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/). It states, "All states updated case and death data daily early in the pandemic, but many have slowed/stopped. As of spring 2022, dozens of states update daily, others provide data once a week. Nebraska and Missouri have stopped providing regular updates".

```{r library, include=FALSE}
library(tidyverse)
library(lubridate)
library(rvest)
library(readxl)
library(tsibble)
library(tqr)
library(scales)

library(maps)
library(mapproj)

library(albersusa)
library(sf)
library(sp)
library(rgeos)
library(maptools)
library(ggalt)
library(ggthemes)
```

```{r read_from_USAfacts, echo=FALSE, cache=FALSE}
# https://usafacts.org/visualizations/coronavirus-covid-19-spread-map/
# Read confirmed cases
conf_raw <- read_csv("https://static.usafacts.org/public/data/covid-19/covid_confirmed_usafacts.csv",
                     col_types = cols(
                       .default = col_double(),
                       `County Name` = col_character(),
                       State = col_character()
                       )
                     )

conf_raw <- conf_raw %>% 
  drop_na(countyFIPS)

conf_total <- conf_raw %>% 
  pivot_longer(!c(1:4), names_to = "publish_date", values_to = "cum_conf") %>% 
  rename(state_abb = State) %>% 
  mutate(publish_date = ymd(publish_date)) %>% 
  group_by(publish_date) %>% 
  summarize(cum_conf = sum(cum_conf), .groups = "drop_last") %>% 
  mutate(
    cum_conf_lag = lag(cum_conf),
    new_conf = cum_conf - cum_conf_lag
  ) %>% 
  select(-cum_conf_lag) %>% 
  mutate(state_abb = "Total")

conf_by_state <- conf_raw %>% 
  pivot_longer(!c(1:4), names_to = "publish_date", values_to = "cum_conf") %>% 
  rename(state_abb = State) %>% 
  mutate(publish_date = ymd(publish_date)) %>% 
  group_by(state_abb, publish_date) %>% 
  summarize(cum_conf = sum(cum_conf), .groups = "drop_last") %>% 
  mutate(
    cum_conf_lag = lag(cum_conf),
    new_conf = cum_conf - cum_conf_lag
  ) %>% 
  select(-cum_conf_lag)

conf <- bind_rows(conf_total, conf_by_state) %>% 
  filter(!is.na(publish_date))

# Read deaths
deaths_raw <- read_csv("https://static.usafacts.org/public/data/covid-19/covid_deaths_usafacts.csv",
                       col_types = cols(
                         .default = col_double(),
                         `County Name` = col_character(),
                         State = col_character()
                         )
                       )

deaths_total <- deaths_raw %>% 
  pivot_longer(!c(1:4), names_to = "publish_date", values_to = "cum_deaths") %>% 
  rename(state_abb = State) %>% 
  mutate(publish_date = ymd(publish_date)) %>% 
  group_by(publish_date) %>% 
  summarize(cum_deaths = sum(cum_deaths), .groups = "drop_last") %>% 
  mutate(
    cum_deaths_lag = lag(cum_deaths),
    new_deaths = cum_deaths - cum_deaths_lag
  ) %>% 
  select(-cum_deaths_lag) %>% 
  mutate(state_abb = "Total")

deaths_by_state <- deaths_raw %>% 
  pivot_longer(!c(1:4), names_to = "publish_date", values_to = "cum_deaths") %>% 
  rename(state_abb = State) %>% 
  mutate(publish_date = ymd(publish_date)) %>% 
  group_by(state_abb, publish_date) %>% 
  summarize(cum_deaths = sum(cum_deaths), .groups = "drop_last") %>% 
  mutate(
    cum_deaths_lag = lag(cum_deaths),
    new_deaths = cum_deaths - cum_deaths_lag
  ) %>% 
  select(-cum_deaths_lag)

deaths <- bind_rows(deaths_total, deaths_by_state) %>% 
  filter(!is.na(publish_date))

# Bind
data_usa <- left_join(conf, deaths, by = c("state_abb", "publish_date"))

# Add state name column
states <- 
  tibble(
    state_abb = c(state.abb, "DC", "Total"),
    state = c(state.name, "District of Columbia", "Total")
  )

data_usa <- data_usa %>% 
  left_join(states, by = "state_abb")

```


```{r preparation, echo=FALSE, message=FALSE}
# add party
gov_html <- read_html("https://en.wikipedia.org/wiki/List_of_United_States_governors")

gov_df <- gov_html %>% 
  html_nodes("table") %>% 
  .[[2]] %>% 
  html_table(fill = TRUE)

gov_df <- gov_df[-1, c(1, 5)]

names(gov_df) <- c("state", "party")

gov_df <- gov_df %>% 
  mutate(party = str_sub(party, 1L, 10L))
#  mutate(party = if_else(state == "Alaska", "Republican", party)) # fill blank

dc_df <- gov_html %>% 
  html_nodes("table") %>% 
  .[[4]] %>% 
  html_table(fill = TRUE)

dc_df <- dc_df[-1, c(1, 5)]

names(dc_df) <- c("state", "party")

gov_dc_df <- bind_rows(gov_df, dc_df)

data_usa <- data_usa %>% 
  left_join(gov_dc_df, by = "state")

# add population
pop <- read_excel("data-raw/nst-est2019-01.xlsx", 
    col_names = FALSE, skip = 9)

pop <- pop[1:51, c(1, 13)]

names(pop) <- c("state", "population")

pop <- pop %>% 
  mutate(
    state = str_sub(state, 2L, -1L)
  ) %>% 
  left_join(states, by = "state")

pop <- pop %>% 
  bind_rows(
    pop %>% 
      summarize(
        state = "Total",
        population = sum(population),
        state_abb = "Total"
      )
  )

data_usa <- data_usa %>% 
  left_join(pop %>% select(-state), by = "state_abb")

data_usa_ma7_std <- data_usa %>% 
  select(-state_abb) %>% 
  mutate(state = str_to_lower(state)) %>% 
  as_tsibble(key = c(state, party, population), index = publish_date) %>% 
  tq_ma(7) %>% 
  as_tibble() %>% 
  mutate(across(cum_conf:new_deaths, ~ .x / population * 1e6))

states_map <- map_data("state") %>% 
  as_tibble() %>% 
  left_join(states %>% mutate(state = str_to_lower(state)),
            by = c(region = "state"))

in_usa <- data_usa %>%
  select(-party) %>%
  pivot_longer(!c(publish_date, state, state_abb, population),
               names_to = "concept", values_to = "value") %>% 
  group_by(state, concept) %>% 
  mutate(
    value_ma = moving_average(value, n = 7, na.rm = TRUE),
    value_per1m = value / population * 1e6,
    value_ma_per1m = value_ma / population * 1e6
      ) %>% 
  ungroup()

state_menu <- unique(in_usa$state)

# make a list
usa <- list(
  map_df = states_map,
  area_df = in_usa,
  area_menu = state_menu
)

usethis::use_data(usa, overwrite = TRUE)

```

I separate red or blue by state governors (in case of DC, a mayor), according to [List of United States governors](https://en.wikipedia.org/wiki/List_of_United_States_governors). I download population data in data directory from [Census Bureau page](https://www.census.gov/data/tables/time-series/demo/popest/2010s-state-total.html), and use estimated population as of July 1, 2019.

```{r map, echo=FALSE}
states_map <- usa_sf("laea")

states_party <- gov_dc_df %>% 
  left_join(pop, by = "state") %>% 
  mutate(pop_mil = population / 1000000) %>% 
  select(-population)

states_map %>% 
  left_join(states_party, by = c("iso_3166_2" = "state_abb")) %>% 
  ggplot() +
  geom_sf(aes(fill = party), color = "white") +  
  scale_fill_manual(values = c("blue", "red")) +
  theme_void()
```

## U.S. confirmed cases (new) by red or blue states

I am worrying that too vigorous "reopen economy" movements in the red states in the United States may cause the second wave.

```{r chart1, echo=FALSE, warning=FALSE}
data_usa %>% 
  filter(state_abb != "Total") %>% 
  group_by(publish_date, party) %>% 
  summarize(new_conf = sum(new_conf), .groups = "drop_last") %>% 
  as_tsibble(key = party, index = publish_date) %>% 
  tq_ma(7) %>% 
  ggplot(aes(publish_date, new_conf, color = party)) +
  geom_line(size = 1) +
  scale_color_manual(values = c("blue", "red")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Confirmed cases (new) by red or blue states",
    x = "published date (7 days average)",
    y = NULL,
    caption = str_c("Latest: ", max(data_usa$publish_date)),
    color = NULL
  ) +
  theme(
    plot.title = element_text(size = rel(2)))
```

Populations in millions by red or blue are:

```{r population_by_party, echo=FALSE}
data_usa %>% 
  filter(state_abb != "Total") %>% 
  filter(publish_date == max(publish_date)) %>% 
  group_by(party) %>% 
  summarize(pop_mil = sum(population) / 1000000, .groups = "drop_last")
```


## Fastest spreading states in the last 14 days

Highest "increase_in_14_days",  which is an increase in cumulative confirmed cases in the last 14 days, are:

```{r spreading, echo=FALSE}
states_speed <- data_usa %>% 
  filter(state_abb != "Total") %>% 
  group_by(state_abb) %>% 
  filter(publish_date > max(publish_date) - 14) %>% 
  summarize(increase_in_14_days = sum(new_conf), .groups = "drop_last")

draw_map_usa <- function(df, var) {
  states_map %>%
    left_join(df, by = c(iso_3166_2 = "state_abb")) %>%
    ggplot() +
    geom_sf(aes(fill = {{ var }}), color = "white") +
    scale_fill_gradient2(low = "#559999", mid = "grey90", high = "#BB650B",
                         midpoint = median(df %>% pull({{ var }}))) +
    theme_void()
}

draw_map_usa(states_speed, increase_in_14_days)

```

Above calculation might be unfair to populous states. Highest "increase_std_in_14_days", which is an increase in cumulative confirmed cases in the last 14 days per 1 million population, are:

```{r standardized, echo=FALSE}
states_speed_std <- data_usa %>% 
  filter(state_abb != "Total") %>% 
  group_by(state_abb) %>% 
  filter(publish_date > max(publish_date) - 14) %>% 
  summarize(increase_std_in_14_days = sum(new_conf * 1000000 / population), .groups = "drop_last")

draw_map_usa(states_speed_std, increase_std_in_14_days)

```


## Highest fatality rate states

Among states with more than 10 cumulative deaths, highest "fatality_rate", which is cumulative deaths per 100 cumulative confirmed cases, are:

```{r fatality_rates, echo=FALSE}
states_fatality <- data_usa %>% 
  filter(state_abb != "Total") %>% 
  mutate(fatality_rate = cum_deaths * 100 / cum_conf) %>% 
  filter(!is.na(fatality_rate)) %>% 
  filter(publish_date == max(publish_date))

draw_map_usa(states_fatality, fatality_rate)

```

## Highest deaths per population states

Highest "deaths_per_1m", which is cumulative deaths per 1 million population, are:

```{r deaths_per_population, echo=FALSE}
states_deaths <- data_usa %>% 
  filter(state_abb != "Total") %>% 
  mutate(
    pop_mil = population / 1e6,
    deaths_per_1m = cum_deaths / pop_mil
    ) %>% 
  filter(!is.na(deaths_per_1m)) %>% 
  filter(publish_date == max(publish_date))

draw_map_usa(states_deaths, deaths_per_1m)

```

EOL

